---
title: "Code from: Phenotypic plasticity in visual opsin gene expression: a meta-analysis in teleost fish"
author: "César Bertinetti"
date: "January 23 2025"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

<style type="text/css">

h1.title {
  font-size: 28px;
  color: Black;
  text-align: justify;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Helvetica", Times, serif;
  color: Black;
  text-align: justify;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Helvetica", Times, serif;
  color: Black;
  text-align: justify;
}
</style>

<style>
pre {
  font-size: 14px;
}
</style>

<br/>

The datasets needed to reproduce this analysis are deposited on Bertinetti et al. 2025. Data from: *Phenotypic plasticity in visual opsin gene expression: a meta-analysis in teleost fish* [Dataset] https://doi.org/10.5281/zenodo.14776297

Following R version and packages were used to run this code:

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pak <- c("readr", "stringr", "matrixStats","dplyr","tidyr","plotrix","RColorBrewer",
         "MESS","zoo","car","Hmisc","corrplot","agricolae","nlme","ggplot2","gridExtra",
         "lme4","rsq","ggpubr","partR2", "MASS", "bbmle","visreg","phylobase","phylosignal",
         "ape","metafor","ggdist", "phangorn","phytools","geiger","rotl")
lapply(pak, require, character.only = TRUE)
``` 
```{r, include=TRUE}

R.version$platform
R.version$version.string
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]

## HERE USE setwd() to set your working directory where the data files will be located

```

### Opsin Gene Expression ###

```{r}
genes <- c("sws1","sws2","rh1","rh2","lws") # gene names

## READ OPSIN DATA WITH Mean, standard deviation and samples from each study
opsin <- read.table("Opsin_Data.csv", header=T, sep =",", dec=".",stringsAsFactors = T)

## READ SPECIES ID INCLUDED TO GENERATE PHYLOTREE
taxon <- read.table("Species_ID.csv", header=T, sep =",", dec=".",stringsAsFactors = F,quote="\"")

### CHECK NUMBER OF SPECIES
length(unique(taxon$Species))
```


<br> Estimate standardized mean differences based on:
Viechtbauer, W. (2010). Conducting meta-analyses in R with the metafor package. Journal of Statistical Software, 36(3), 1–48. https://doi.org/10.18637/jss.v036.i03

```{r}

opsin_meta <-  escalc(measure="SMD", m1i=Mean1, sd1i=SD1, n1i=N1,
                      m2i=Mean2, sd2i=SD2, n2i=N2, di=Dval, data=opsin,slab = Author)

##  READ GENERATED PHYLOGENETIC TREE
tree <- read.tree("plast.tre")
### Transform tree into an ultrametric one
tree <- compute.brlen(tree)
### compute phylogenetic correlation matrix
A <- vcv(tree, corr=TRUE)

### MERGED PHYLOGENTIC TREE AND OPSIN PLASTICITY DATA
opsin_master <- merge(opsin_meta, taxon[,-c(2:3,5)], by="Species",all=T)
opsin_master$search_string <- sub(" ", "_", opsin_master$search_string)
colnames(opsin_master)[23] <- "Tree_ID"
# sum(is.na(opsin_master$Tree_ID)) ### Check no species IDs are mising values!

### USE ABSOLUTE SMD AS A MEASURE OF PLASTICITY REGARDLESS OF DIRECTION OF CHANGE
opsin_master$yi <- abs(opsin_master$yi)
```

<br> Run mixed-effect meta-analysis models including moderators and random effects.

```{r}
model = ~ Treatment * Type 
## model = ~ 0 + Treatment:Type can be used to leave out model intercept

## the following model takes a while to run
## multi-level phylogenetic meta-analytic model
overall_int <- rma.mv(yi, vi,
                      mods = model,
                      random = list(~ 1 | Study/Sample, ~ 1 | Species, ~ 1 | Tree_ID),
                      R=list(Tree_ID=A), data=opsin_master, control=list(rel.tol=1e-8))
overall_int
anova(overall_int, btt=3:4) ## test interaction among moderators
```

<br> This can now be redone to test heterogeneity depending on experimental treatments or moderators. Requires creating smaller phylogenetic trees for eachd dataset. 

```{r}
## LIGHT referred as External Stimuli
kp_sp <- unique(subset(opsin_master, Treatment=="light")$Tree_ID)

over_phy <- vcv(compute.brlen(drop.tip(tree, setdiff(tree$tip.label, kp_sp))), corr=T) 

## Run model, takes some time
overall_l <- rma.mv(yi, vi, subset=(Treatment=="light"),
                     random = list(~ 1 | Study/Sample, ~ 1 | Species, ~ 1 | Tree_ID),
                     R=list(Tree_ID=over_phy), data=opsin_master, control=list(rel.tol=1e-8),slab = paste(Author, Sample, Age_dph, sep=", "))

overall_l



```

<br> This can be expanded to test moderator (e.g, light spectra) effect on each opsin.

```{r}
mod_light_opsin = ~ 0 + Type:Opsin_Class
overall_l_ops_mod <- rma.mv(yi, vi, subset=(Treatment=="light"), mods=mod_light_opsin,
                        random = list(~ 1 | Study/Sample, ~ 1 | Species, ~ 1 | Tree_ID),
                        R=list(Tree_ID=over_phy), data=opsin_master, control=list(rel.tol=1e-8),slab = paste(Author, Sample, Age_dph, sep=", "))
overall_l_ops_mod

anova(overall_l_ops_mod, X=c(1,-1,0,0,0,0,0,0,0,0)) # for example for lws red-sensitive opsin
```

<br> Or this can also be used to test overall effect sized on each opsin

```{r}
sws1 <- subset(opsin_master, Opsin_Class == "sws1") # Subset opsin of interest
## Drop unneeded species by selecting the ones present in the subset
kp_sp <- unique(subset(opsin_master, Opsin_Class=="sws1")$Tree_ID)
sws1_phy <- vcv(compute.brlen(drop.tip(tree, setdiff(tree$tip.label, kp_sp))), corr=T) 

## Run model - Overall
res.sws1 <- rma.mv(yi, vi,
                   random = list(~ 1 | Study/Sample, ~ 1 | Species, ~ 1 | Tree_ID),
                   R=list(Tree_ID=sws1_phy), data=sws1, control=list(rel.tol=1e-8),slab = paste(Author, Sample, Age_dph, sep=", "))

res.sws1
```

## PHYLOGENETIC SIGNAL ## 

<br> This mostly based on the phylosignal package:

Keck F., Rimet F., Bouchez A. & Franc A. (2016) phylosignal: an R package to measure, test, and explore the phylogenetic signal. Ecology and Evolution 6 (9), 2774–2780. http://onlinelibrary.wiley.com/doi/10.1002/ece3.2051/epdf

```{r}
## JUST A BASIC FUNCTION
momdis = function(x) {
  c(m = mean(x), s = sd(x), iqr = IQR(x), n = length(x), se = std.error(x), max = max(x))
}

### SUMMARISE PLASTICITY VALUES FOR EACH SPECIES
mean_plast <- aggregate(yi ~ Species + Tree_ID, data = opsin_master, momdis)
mean_plast <- do.call(data.frame, mean_plast)

p4 <-  data.frame(mean_plast[,3])
row.names(p4) <- mean_plast$Tree_ID

## COMBINE TREE AND MEAN PLASTICITY
p4d <- phylo4d(tree, p4) 

#Assessing the signal depth with correlograms
plast.crlg <- phyloCorrelogram(p4d, trait = "mean_plast...3.")
plot(plast.crlg)
```





